---
title: "Tidyverse Practice Problem Solutions"
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
---

1. Import the excel file `data/titanic3.xls` and select the `age` column.

```{r}
# Load the package to read excel files
library(tidyverse)
library(readxl)

# Upload the data and assign it to an object
titanic <- read_excel(path = "data/titanic3.xls")

# Select the age column
titanic$age

# or

titanic %>% select(age)
```

2. In the iris dataset what is the average Sepal.Length per species?

```{r}
iris %>% as_tibble() %>% group_by(Species) %>% summarise(Sepal.Length.avg = mean(Sepal.Length))
```


3. How many passengers on the titantic were older than 20?

```{r}
titanic %>% filter(age > 30) %>% count()
```

4. What is the average age of men and women on the titanic?
```{r}
titanic %>% select(age) %>% drop_na() %>% summarise(mean(age))
```

5. How many pasengers survived the sinking of the titanic?
```{r}
titanic %>% filter(survived == 1) %>% count()

# ...and sadly passed away
titanic %>% filter(survived == 0) %>% count()
```

6. What age was the oldest person to survive the sinking of the titanic?
```{r}
titanic %>% filter(survived == 1) %>% arrange(desc(age))
```

7. From the iris data set, find the average Sepal.Length per species then export the resulting data frame as a .tsv file in the `/data/` directory of the RProject
```{r}
iris %>% group_by(Species) %>% summarise(Sepal.Length.avg = mean(Sepal.Length))
```
